---
title: "Spatial Statistics Learning Group"
subtitle: "Regression modelling for spatially referenced data<html><div style='float:left'></div><hr color='#324147' size=4px width=800px></html>"
date: "<br/>Week 2: 6 November 2019"
output:
  xaringan::moon_reader:
    css: ["default", "mtheme.css", "fonts_mtheme.css", "my-theme.css"]
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
editor_options: 
  chunk_output_type: console
---
```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
```
class: inverse, middle, center

# A review of generalized linear models

---
# Generalized linear models

**Assumptions:**

- $Y_i\sim f(\cdot)$ belongs to the exponential family (i.e. Normal, Poisson, Binomial...) 

--

- $E[Y_i]=m_i\mu_i$ and $Var[Y_i]=m_iV(\mu_i)$

--

- $g(\mu_i)=\eta_i=d_{i}^{\text{T}}\beta$

--

- $Y_i$ are mutually independent for $i=1,\ldots,n$

---
# Overdispersion

**Definition:** the data show a greater variability than that implied by a classical GLM.

--

Possible cause of over dispersion are:
- omission of risk factors
- intrinsic random variation that results from individual characteristics
- **spatial correlation**

--

$$Var[Y_i + Y_j]=Var[Y_i]+Var[Y_j]+2Cov(Y_i, Y_j)$$

--

Solution for the first two:
- *random effects models*: $\eta_i=d_{i}^{\text{T}}\beta+Z_i$ where $Z_i$ is a set of *independent* Gaussian variables with mean zero and variance $\sigma^2$. 
---
# Generalized linear mixed models
**Assumptions:**

- $Z_i$ are i.i.d. random variables

--

- $Y_i\mid Z_i\sim f(\cdot)$ belongs to the exponential family 

--

- $E[Y_i\mid Z_i]=m_i\mu_i$ and $Var[Y_i\mid Z_i]=m_iV(\mu_i)$

--

- $g(\mu_i)=\eta_i=d_{i}^{\text{T}}\beta+Z_i$

--

- $Y_i\mid Z_i$ are mutually independent for $i=1,\ldots,n$

--

In this way we can account for extra non-spatial variability in the data. This will be our starting point to look for evidence of spatial dependence in the data.
---
class: inverse, middle, center

# Questioning the assumption of independence

---
# Independence

One of the main assumptions of generalised linear (mixed) models is tha the observations $Y_i$ are mutually independent. 

--

.center[**Is this assumption is still valid for spatial data?**]

--

**First law of geography:** close things are more related than distant things.

--

There are two main consequences for violating this assumptions:

1. Narrower confidence intervals for the regression parameters that leads to an increase in Type I error.
2. We don't exploit the spatial correlation when doing predictions. 

---
class: center, middle, inverse

# Testing for spatial correlation

---
# The theoretical variogram

It measures the similarity of values as a function of the distance between their locations. It is an exploratory tool that helps to asses the presence of spatial dependece in the data. 

--

The **theoretical variogram** is defined as:
$$V_{Y}(u_{ij})=\frac{1}{2}E\left\{[Y(x_{i})-Y(x_{j})]^2\right\}$$
where $u_{ij}$ is the euclidean distance between point $x_i$ and point $x_j$

--

Suppose random variables $Y(x_i)$ and $Y(x_j)$ have expectation zero, variance $\sigma^2$ and correlation $\rho(u_{ij})$ then

$$V_{Y}(u_{ij})=\sigma^2\{1 - \rho(u_{ij})\}$$
---
#What is the variogram estimating?

$$V_{Y}(u)=\sigma^2\{1 - \rho(u)\}$$
The function $\rho(u)$ is called the **spatial correlation function**, is the statistical version of Tobler's law (decreases with increasing $u$)

--

**Key assumption is stationarity:**
- spatially constant expectation, $\mu(x)=\mu$
- spatially constant variance, $\sigma^2(x)=\sigma^2$
- correlation is a function of distance

--

**How to accommodate explanatory variables?**
- fit a provisional non-spatial regression model
- calculate variogram of standardized residuals
---
# The theoretical variogram
```{r semivario, echo=FALSE, fig.align='center', fig.width=9, fig.height=6, dpi=600, warning=FALSE, message=FALSE}
library(geoR)
kappa = 0.5
model = "matern"
phi = 0.25
sigma= 1
nugget = 0
prange = practicalRange(model, phi, kappa)

u <- seq(0,1.5, l = 500)
plot(u, nugget + sigma - cov.spatial(u, model, cov.pars = c(sigma,phi),kappa), xlim = c(0,1.5), ylim = c(0,1.2), type = "l", ylab = expression(V[Y](u)), col = "darkblue", lwd = 3)

segments(0, nugget + sigma, 1.5, nugget + sigma, col = "darkblue", lwd = 1.5, lty = 2)
#segments(0, nugget, 1.5, nugget, col = "darkblue", lwd = 1.5, lty = 2)

segments(prange, 0.95, prange, 0,col = "darkblue", lty = 2, lwd = 1.5)

text(x = 0.05, y = 1.05, label = expression(sigma^2~"="~"sill") , col = "darkblue")
text(x = 0.75, y = 0, label = "practical range" , col = "darkblue")
```
---
# .font80[The cloud and empirical semivariogram]

The obseserved quantities
$$v_{ij}=\frac{1}{2}(y_{i}-y_{j})^2$$
constitute the **cloud semivariogram**.  

If then we average values of $v_{ij}$ over suitably narrow ranges of inter-point distances $u_{ij}$ we obtain the **empirical semivariogram**


```{r, echo=F, fig.align="center", fig.width=12, fig.height=5, dpi=600}
library(geoR)
v1 <- variog(elevation, messages = F, op = "cloud")
v2 <- variog(elevation, messages = F)

par(mfrow = c(1, 2))
plot(v1, main = "Cloud Semivariogram")
abline(v = v2$u, col = "red", lty = 2)
plot(v2, main = "Empirical Semivariogram")
```

---
# Testing for spatial correlation

How do we know if the sample variogram shows evidence of spatial correlation?

1. Permute the locations $x_1,\ldots,x_n$ while holding fix the rest of the data.
2. Compute the sample variogram for the permuted data.
3. Repeat 1 and 2 a large enough number of times (e.g. 1000).
4. Compute the 95% confidence intervals at each binned distance $u$.

**If the sample variogram falls entirely within the 95% confidence intervals then there is NO evidence of spatial correlation.** 

**Note:** it is always a good idea to conduct this test on the residuals of a non-spatial model rather than on the observed outcome.